---
title: "Bio Inv Revisions"
author: "Caitlin Mothes, PhD"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(terra)
library(sf)
library(mapview)
library(spThin)
library(spdep)
library(adegenet)
```

# Read in Data

```{r}
## raw 10m landcover data
lc <- rast("data/landcover_mdc.tif")

## reclassify based on crosswalk doc

reclass <- read_csv("data/CLC_to_FNAINC_Crosswalk_20160701.csv")


land_reclass <- terra::classify(lc, as.matrix(reclass %>% dplyr::select(SITE, reclass)))
# values 2134 and 5300 in raster but not in table...
land_reclass[land_reclass == 5300] <- 16
land_reclass[land_reclass == 2134] <- 10

## read in occurrences

## COMBINE AND REMOVE DUPLICATES
### Agama
ag_all <- read_csv("data/agama_occ_mdc.csv") %>% select(LONG = Longitude, LAT = Latitude)

agama_pts <- read_csv("data/agama_sites.csv") %>% select(LONG, LAT) 

ag_combined <- bind_rows(ag_all, agama_pts) %>% distinct(LONG, LAT) %>% 
  mutate(species = "Agama")


### Basilisk
bas_all <- read_csv("data/basilisk_occ_mdc.csv") %>% select(LONG = Longitude, LAT = Latitude)

bas_pts <- read_csv("data/basilisk_sites.csv") %>% select(LONG, LAT)

bas_combined <- bind_rows(bas_all, bas_pts) %>% distinct(LONG, LAT) %>% 
 mutate(species = "Basilisk")

#mapview::mapview(bas_combined)

```

# Agama Land Cover Association

## Thin Points and clean

```{r}
## thin points (increasing thin.par as needed)
ag_thin <- thin(ag_combined, lat.col = "LAT", long.col = "LONG", spec.col = "species",
                thin.par = 0.2,
                reps = 50, write.files = FALSE, locs.thinned.list.return = TRUE)


## test for autocorrelation with join counts test, using last df
ag_thin_sf <- ag_thin[[1]] %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>% 
  st_transform(crs = st_crs(land_reclass))
```

## Set up for Join Counts test

```{r}
## create raster for test
r_ag <- rast(extent = ext(ag_thin_sf), resolution = 100, crs = crs(land_reclass))

r_ag2 <- rasterize(ag_thin_sf, r_ag, field = 1)
# make NA values 0
r_ag2[is.na(r_ag2)] <- 0

# check how many points were lost to rasterization at slightly larger res
freq(r_ag2) # two, n = 406
```

## Run Join Counts Tests

```{r}
## set up connectivity matrix, rook default
nb <- cell2nb(nrow = nrow(r_ag2), ncol = ncol(r_ag2))

## convert to w weights list using Binary weights matrix
lwb <- nb2listw(nb, style = "B")

#join counts test and permuation test
ag_jc <- joincount.test(as.factor(values(r_ag2)), lwb, alternative = "greater")

ag_jc_perm <- joincount.mc(as.factor(values(r_ag2)), lwb, nsim = 999, alternative = "greater")


ag_jc
# 	Join count test under nonfree sampling
# 
# data:  as.factor(values(r_ag2)) 
# weights: lwb 
# 
# Std. deviate for 0 = -0.14703, p-value = 0.5584
# alternative hypothesis: greater
# sample estimates:
# Same colour statistic           Expectation              Variance 
#          8.537650e+05          8.537653e+05          3.248657e+00 
# 
# 
# 	Join count test under nonfree sampling
# 
# data:  as.factor(values(r_ag2)) 
# weights: lwb 
# 
# Std. deviate for 1 = -0.87636, p-value = 0.8096
# alternative hypothesis: greater
# sample estimates:
# Same colour statistic           Expectation              Variance 
#             0.0000000             0.7665537             0.7651036 

ag_jc_perm
# Monte-Carlo simulation of join-count statistic
# 
# data:  as.factor(values(r_ag2)) 
# weights: lwb 
# number of simulations + 1: 1000 
# 
# Join-count statistic for 0 = 853765, rank of observed statistic = 485.5, p-value = 0.5145
# alternative hypothesis: greater
# sample estimates:
#     mean of simulation variance of simulation 
#           8.537652e+05           2.909942e+00 
# 
# 
# 	Monte-Carlo simulation of join-count statistic
# 
# data:  as.factor(values(r_ag2)) 
# weights: lwb 
# number of simulations + 1: 1000 
# 
# Join-count statistic for 1 = 0, rank of observed statistic = 235.5, p-value = 0.7645
# alternative hypothesis: greater
# sample estimates:
#     mean of simulation variance of simulation 
#              0.7717718              0.7634909 
```

## Recalculate LC Metrics

```{r}
ag_lc <- ag_thin_sf %>% 
  mutate(landcover = as.numeric(extract(lc, .)[[2]])) %>%
  left_join(reclass, by = c("landcover" = "SITE")) %>% 
  st_drop_geometry() %>% 
  count(NAME_STATE) %>% 
  # these calculations from Table 3 Metrics spreadsheet (from Chris)
  mutate(percent = n/sum(n)*100,
         point_estimate = (n+2)/(sum(n)+4),
         L95 = point_estimate-1.96*sqrt((point_estimate*(1-point_estimate))/(sum(n)+4)),
         U95 = point_estimate+1.96*sqrt((point_estimate*(1-point_estimate))/(sum(n)+4)),
         # calculate 83% intervals to appease reviewer
         L83 = point_estimate-1.38*sqrt((point_estimate*(1-point_estimate))/(sum(n)+4)),
         U83 = point_estimate+1.38*sqrt((point_estimate*(1-point_estimate))/(sum(n)+4)),
         species = "Agama")


```

# Basilisk Land Cover Association

## Thin points and clean

```{r}
## thin points (increasing thin.par as needed)
bas_thin <- thin(bas_combined, lat.col = "LAT", long.col = "LONG", spec.col = "species",
                thin.par = 0.15,
                reps = 50, write.files = FALSE, locs.thinned.list.return = TRUE)


## test for autocorrelation with join counts test, using last df
bas_thin_sf <- bas_thin[[1]] %>%
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326) %>% 
  st_transform(crs = st_crs(land_reclass))

```

## Set up for Join Counts Test

```{r}
## create raster for test
r_bas <- rast(extent = ext(bas_thin_sf), resolution = 100, crs = crs(land_reclass))

r_bas2 <- rasterize(bas_thin_sf, r_bas, field = 1)
# make NA values 0
r_bas2[is.na(r_bas2)] <- 0

# check how many points were lost to rasterization at slightly larger res
freq(r_bas2) # only 3, n = 311

```

## Run Join Counts Test

```{r}
## set up connectivity matrix, rook default
nb <- cell2nb(nrow = nrow(r_bas2), ncol = ncol(r_bas2))

## convert to w weights list using Binary weights matrix
lwb <- nb2listw(nb, style = "B")

#join counts test and permuation test
bas_jc <- joincount.test(as.factor(values(r_bas2)), lwb, alternative = "greater")

bas_jc_perm <- joincount.mc(as.factor(values(r_bas2)), lwb, nsim = 999, alternative = "greater")


bas_jc
# Join count test under nonfree sampling
# 
# data:  as.factor(values(r_bas2)) 
# weights: lwb 
# 
# Std. deviate for 0 = 1.2204, p-value = 0.1111
# alternative hypothesis: greater
# sample estimates:
# Same colour statistic           Expectation              Variance 
#          6.307360e+05          6.307339e+05          2.903259e+00 
# 
# 
# 	Join count test under nonfree sampling
# 
# data:  as.factor(values(r_bas2)) 
# weights: lwb 
# 
# Std. deviate for 1 = 1.7871, p-value = 0.03696
# alternative hypothesis: greater
# sample estimates:
# Same colour statistic           Expectation              Variance 
#             2.0000000             0.6079493             0.6067576 

bas_jc_perm
# Monte-Carlo simulation of join-count statistic
# 
# data:  as.factor(values(r_bas2)) 
# weights: lwb 
# number of simulations + 1: 1000 
# 
# Join-count statistic for 0 = 630736, rank of observed statistic = 864.5, p-value = 0.1355
# alternative hypothesis: greater
# sample estimates:
#     mean of simulation variance of simulation 
#           6.307339e+05           2.969003e+00 
# 
# 
# 	Monte-Carlo simulation of join-count statistic
# 
# data:  as.factor(values(r_bas2)) 
# weights: lwb 
# number of simulations + 1: 1000 
# 
# Join-count statistic for 1 = 2, rank of observed statistic = 916.5, p-value = 0.0835
# alternative hypothesis: greater
# sample estimates:
#     mean of simulation variance of simulation 
#              0.6396396              0.6275253 

```

## Recalc LC Metrics

```{r}
bas_lc <- bas_thin_sf %>% 
  mutate(landcover = as.numeric(extract(lc, .)[[2]])) %>%
  left_join(reclass, by = c("landcover" = "SITE")) %>% 
  st_drop_geometry() %>% 
  count(NAME_STATE) %>% 
  mutate(percent = n/sum(n)*100,
         point_estimate = (n+2)/(sum(n)+4),
         L95 = point_estimate-1.96*sqrt((point_estimate*(1-point_estimate))/(sum(n)+4)),
         U95 = point_estimate+1.96*sqrt((point_estimate*(1-point_estimate))/(sum(n)+4)),
         # calculate 83% intervals to appease reviewer
         L83 = point_estimate-1.38*sqrt((point_estimate*(1-point_estimate))/(sum(n)+4)),
         U83 = point_estimate+1.38*sqrt((point_estimate*(1-point_estimate))/(sum(n)+4)),
         species = "Basilisk")




```

# Contingency Analysis

### Join datasets

```{r}
land_assoc <- bind_rows(ag_lc, bas_lc) %>% 
  group_by(NAME_STATE) %>% 
  mutate(sum_obs = sum(n)) %>% 
  filter(sum_obs > 10)  %>% 
  # remove marshes and terrestrial, both less than 10
  filter(!NAME_STATE %in% c("Cultural - Terrestrial", "Marshes"))
  # filter to categories that have more than 10 for at least one species
  # group_by(NAME_STATE) %>% 
  # mutate(n_cats = n()) %>% 
  # filter(n_cats > 1) %>% 
  # ungroup() %>% 
  # select(-n_cats)
  
```

### Contingency Analysis

```{r}

full_join(bas_lc, ag_lc, by = "NAME_STATE") %>%
  rename(bas_n = n.x, ag_n = n.y) %>%
  replace(is.na(.), 0) %>%
  #remove cover types with < 10 occ for both species
  filter(bas_n >= 10 | ag_n >= 10) %>% 
  column_to_rownames(var = "NAME_STATE") %>% select(bas_n, ag_n) %>% 
  chisq.test(.) 

#X-squared = 74.387, df = 8, p-value = 6.542e-13

 
```

### CI Analysis and Visual

```{r}
## confidence interval analysis

  
#assess CI overlap
land_assoc %>%
  # Add asterisk to sig diff. based on 83% CI
  mutate(
    NAME_STATE = if_else(
      !NAME_STATE %in% c("Transportation", "Low Intensity Urban"),
      paste0(NAME_STATE, "*"),
      NAME_STATE
    ),
    # rename cultural - riverine to Canal/Ditch)
    NAME_STATE = case_when(str_detect(NAME_STATE, "Riverine") ~ "Canal/Ditch*", 
                           str_detect(NAME_STATE, "Cropland") ~ "Cropland/ Pasture*",
                           str_detect(NAME_STATE, "Prairies") ~ "Prairies/ Bogs*",
                           str_detect(NAME_STATE, "Nurseries") ~ "Vineyard/ Nurseries*",
                           .default = NAME_STATE)
  ) %>%
  ggplot(aes(x = species, y = point_estimate, color = species)) +
  geom_point(size = 6) +
  geom_errorbar(aes(ymin = L95, ymax = U95)) +
  scale_color_manual(name = "Species", values = c("blue", "red")) +
  ylab("Probability of Occurence") +
  theme_bw() +
  facet_wrap(~ NAME_STATE, nrow = 1, strip.position = "bottom", labeller = labeller(NAME_STATE = label_wrap_gen(10))) +
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    axis.ticks.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.major.x = element_blank(),
    text = element_text(size = 18),
    legend.position = "bottom"
  )
```

# Minor Rev -Landcover Association

Reviewer #3 suggested comparing habitat types of all samples (sites visited) compared to all identified sites (GBIF etc. localities) with a chi square test

### Agamas

```{r}
# _pts = samples, _all = gbif etc.

#land cover associate for both
ag_pts_lc <- agama_pts %>% 
  st_as_sf(coords = c("LONG", "LAT"), crs = 4326) %>% 
  st_transform(crs = st_crs(land_reclass)) %>% 
  mutate(landcover = as.numeric(extract(lc, .)[[2]])) %>%
  left_join(reclass, by = c("landcover" = "SITE")) %>% 
  st_drop_geometry() %>% 
  count(NAME_STATE)

ag_all_lc <- ag_all %>% 
  st_as_sf(coords = c("LONG", "LAT"), crs = 4326) %>% 
  st_transform(crs = st_crs(land_reclass)) %>% 
  mutate(landcover = as.numeric(extract(lc, .)[[2]])) %>%
  left_join(reclass, by = c("landcover" = "SITE")) %>% 
  st_drop_geometry() %>% 
  count(NAME_STATE)

# chi square test
ag_all_lc %>% 
  rename(all_sites = n) %>% 
  # try just categories in sampled sites
  full_join(ag_pts_lc, by = "NAME_STATE") %>% 
  #right_join(ag_pts_lc, by = "NAME_STATE") %>%
  rename(sampled_sites = n) %>% 
  mutate(across(where(is.numeric), ~replace_na(.x, 0))) %>% 
    # remove categories with less than 5 for both
  filter(sampled_sites >=5 | all_sites >= 5) %>% 
  column_to_rownames("NAME_STATE") %>% 
  data.matrix() %>% 
  chisq.test()

#significant...but I think that is due to huge sample size differences? (for all cats and just sampled cats)
## When removing categories that had less than 5 obs for both groups difference is insignificant

# try just comparing percentages
ag_all_lc %>% 
  rename(all_sites = n) %>% 
  full_join(ag_pts_lc, by = "NAME_STATE") %>% 
  #right_join(ag_pts_lc, by = "NAME_STATE") %>% 
  rename(sampled_sites = n) %>% 
  mutate(across(where(is.numeric), ~replace_na(.x, 0))) %>% 
  mutate(all_sites_pcnt = all_sites/sum(all_sites)*100,
         sampled_sites_pcnt = sampled_sites/sum(sampled_sites)*100) %>% 
  select(-c(2:3)) %>% 
  column_to_rownames("NAME_STATE") %>% 
  data.matrix() %>% 
  chisq.test()

# no difference in percentage of sites in each lc category (for both full and just sampled categories)

```

### Basilisks

```{r}
# _pts = samples, _all = gbif etc.

#land cover associate for both
bas_pts_lc <- bas_pts %>% 
  st_as_sf(coords = c("LONG", "LAT"), crs = 4326) %>% 
  st_transform(crs = st_crs(land_reclass)) %>% 
  mutate(landcover = as.numeric(extract(lc, .)[[2]])) %>%
  left_join(reclass, by = c("landcover" = "SITE")) %>% 
  st_drop_geometry() %>% 
  count(NAME_STATE)

bas_all_lc <- bas_all %>% 
  st_as_sf(coords = c("LONG", "LAT"), crs = 4326) %>% 
  st_transform(crs = st_crs(land_reclass)) %>% 
  mutate(landcover = as.numeric(extract(lc, .)[[2]])) %>%
  left_join(reclass, by = c("landcover" = "SITE")) %>% 
  st_drop_geometry() %>% 
  count(NAME_STATE)

# chi square test
bas_all_lc %>% 
  rename(all_sites = n) %>% 
  # try just categories in sampled sites
  full_join(ag_pts_lc, by = "NAME_STATE") %>% 
  #right_join(bas_pts_lc, by = "NAME_STATE") %>% 
  rename(sampled_sites = n) %>% 
  mutate(across(where(is.numeric), ~replace_na(.x, 0))) %>% 
     # remove categories with less than 5 for both
  filter(sampled_sites >=5 | all_sites >= 5) %>% 
  column_to_rownames("NAME_STATE") %>% 
  data.matrix() %>% 
  chisq.test()

#different only when considering all sites (even when remove small sample size categories), but not for just sampled sites

# try just comparing percentages
bas_all_lc %>% 
  rename(all_sites = n) %>% 
  full_join(ag_pts_lc, by = "NAME_STATE") %>% 
  #right_join(bas_pts_lc, by = "NAME_STATE") %>% 
  rename(sampled_sites = n) %>% 
  mutate(across(where(is.numeric), ~replace_na(.x, 0))) %>% 
  mutate(all_sites_pcnt = all_sites/sum(all_sites)*100,
         sampled_sites_pcnt = sampled_sites/sum(sampled_sites)*100) %>% 
  select(-c(2:3)) %>% 
  column_to_rownames("NAME_STATE") %>% 
  data.matrix() %>% 
  chisq.test()

# no difference in percentage of sites in each lc category for just sampled categories, but different when considering all categories
```

# Figure 1 updates

```{r}
library(tmap)
library(basemaps)
# create basemap of urbanization and make points smaller and transparent

# combine basilisk and agama points
sites <- bas_pts %>% mutate(species = "Basilisk") %>% 
  bind_rows(mutate(agama_pts, species = "Agama")) %>% 
  #remove basilisk outlier
  slice(-48) %>% 
  st_as_sf(coords = c("LONG", "LAT"), crs = 4326)

mapview(sites)

#create raster of urbanization
urbanization <- land_reclass
urbanization[urbanization == 4] <- 100
urbanization[urbanization == 6] <- 200
urbanization[!urbanization %in% c(100,200)] <- NA

qtm(urbanization)

# read in MDC layer
mdc <- read_sf("data/MDC_land_boundary.shp")

# set basemap
set_defaults(map_service = "esri", map_type = "world_imagery")
bgbox <-  st_bbox(st_buffer(sites, 5000)) %>% st_as_sfc()
bg <- basemap_stars(bgbox) #process basemap into a raster

tm_shape(bg) +
  tm_rgb() +
  tm_shape(mdc) +
  tm_borders() +
  #tm_shape(urbanization) +
  # tm_raster(
  #   style = "cat",
  #   palette = c("orange", "#f7ef4f"),
  #   labels = c("High Intensity Urban", "Low Intensity Urban"),
  #   alpha = 0.75
  # ) +
  tm_shape(sites) +
  tm_dots(
    shape = "species",
    col = "species",
    border.col = "black",
    shapes = c(24, 22),
    palette = c("red", "blue"),
    size = 1,
    alpha = 0.75
    #legend.show = FALSE
  ) +
  tm_layout(legend.show = FALSE) +
  tm_compass(type = "arrow", position = c(.90, .05), text.color = "white") +
  tm_scale_bar(position = c(0.58, .005), text.size = .8, text.color = "white")

```

# sPCA Figure Updates

```{r}
load("data/agama_spca2eig.RData")
load("data/agama_genind.RData")

colorplot.spca(mySpca, transp = TRUE)

showBauges()
colorplot(mySpca$other$xy, rupica.spca1$ls, axes=1:2, transp=TRUE, add=TRUE,
cex=3)
title("sPCA - colorplot of PC 1 and 2\n(lagged scores)", col.main="yellow",
      line=-2, cex=2) 

```

# Figure 7 Updates

```{r}
# https://rpubs.com/camerynbrock/963528 

library(rnaturalearth) 

# read in resistance layer
canopy <- terra::rast("data/canopy.asc")
crs(canopy) <- crs(land_reclass)

# read in basemap
basemap_sf <- ne_countries(
  continent = "North America",
  scale = "medium", 
  returnclass = "sf") %>% 
  dplyr::select(1) %>% 
  st_transform(crs = crs(canpoy)) 

colors <- colorRampPalette(
  colors = c("#85c1c8", "#af4980", "#c0182a", "#d33300", "#e99900", 
             "#ffff00"),
  bias = 0.25) 

plot(canopy, col = colors(100)) 

tm_shape(canopy) +
  tm_raster(style = "cont") +
  tm_shape(filter(sites, species == "Basilisk") %>% st_transform(crs = crs(canopy))) +
  tm_dots(size = 0.5, col = "blue", border.col = "white") +
  tm_layout(legend.outside = TRUE,
            frame = FALSE)
  
 
```

# Intro History Map Updates

Add original basilisk site to 197# map

```{r}
# read in intro sites
intro_sites <- readxl::read_xlsx("data/introduction_sites.xlsx", sheet = 1) %>% 
  select(Longitude, Latitude, year = Year, Species)


# AGama ------------

# get sites w/ year info
ag_history <- read_csv("data/agama_occ_mdc.csv") %>% 
  select(year:Latitude) %>% 
  bind_rows(filter(intro_sites, Species == "Agama")) %>% 
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

#plot specific range of years
years <- unique(ag_history$year) %>% sort()

p <- vector("list", length = length(years))
for (i in 1:length(years)) {
  p[[i]] <- ggplot() +
    geom_sf(data = mdc,
                 fill = "lightgrey") +
    theme(
      panel.background = element_blank(),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      plot.margin = unit(c(0,01,0.1,0), "lines")
    ) +
    geom_sf(
      data = ag_history[ag_history$year <= years[i], ],
      # aes(Longitude, Latitude),
      # inherit.aes = FALSE,
      shape = 15,
      colour = "blue"
    )
}

library(ggpubr)
#arrange all together
ggarrange(p[[1]], p[[2]], p[[3]], p[[4]], p[[5]], p[[6]], p[[7]],
          p[[8]], p[[9]], p[[10]], p[[11]], p[[12]], p[[13]], p[[14]],
          p[[15]], p[[16]], p[[17]], p[[18]], p[[19]], p[[20]],
          nrow = 5, ncol = 4, labels = years[1:20])


## basilisk ---------
# get sites w/ year info
bas_history <- read_csv("data/basilisk_occ_mdc.csv") %>% 
  select(year:Latitude) %>% 
  bind_rows(filter(intro_sites, Species == "Basilisk")) %>% 
  st_as_sf(coords = c("Longitude", "Latitude"), crs = 4326)

#plot specific range of years
years <- unique(bas_history$year) %>% sort()

p <- vector("list", length = length(years))
for (i in 1:length(years)) {
  p[[i]] <- ggplot() +
    geom_sf(data = mdc,
                 fill = "lightgrey") +
    theme(
      panel.background = element_blank(),
      axis.title = element_blank(),
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      plot.margin = unit(c(0,01,0.1,0), "lines")
    ) +
    geom_sf(
      data = bas_history[bas_history$year <= years[i], ],
      # aes(Longitude, Latitude),
      # inherit.aes = FALSE,
      shape = 15,
      colour = "red"
    )
}

#arrange all together
ggarrange(p[[1]], p[[2]], p[[3]], p[[4]], p[[5]], p[[6]], p[[7]],
          p[[8]], p[[9]], p[[10]], p[[11]], p[[12]], p[[13]], p[[14]],
          p[[15]], p[[16]], p[[17]], p[[18]], p[[19]], p[[20]],
          nrow = 5, ncol = 4, labels = years[1:20])

```

# Kinship

```{r}
ag_kinship <- read_csv("data/agama_kinshipPlink.csv") %>% 
  select(-1) %>% 
  as.matrix() 

max(ag_kinship[lower.tri(ag_kinship)]) # row 13, col 41 
# read in site metadata
site_meta <- read_csv("data/collection_sites_updated.csv") 

# map the ones with the highest kinship
site_meta %>% 
  slice(c(13, 41)) %>% 
  st_as_sf(coords = c("LONG", "LAT"), crs = 4326) %>% 
  mapview()
```
